---
title: "Identifing Environmental Events With Greatest Health and Economic Effects"
author: "Philip Tabak"
date: "`r Sys.Date()`"
output: html_document
keep_md: yes
---

Unzip and read source data. We will also load libraries that we will use later in analysis at this stage.


```{r initialize}

library(dplyr)
library(qdap)

source_url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
local_filename = "StormData.csv.bz2"

writeLines(capture.output(sessionInfo()), "sessionInfo.txt")

```


##Synopsis
###Does the document have a synopsis that describes and summarizes the data analysis in less than 10 sentences?


This analysis is preformed on the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. Data was downloaded on `r Sys.Date()` from `r source_url`.



##Data Processing
###Is there a section titled "Data Processing" that describes how the data were loaded into R and processed for analysis?


```{r get-and-clean-source-data}


download.file(source_url, destfile = local_filename)

storm_data <- read.csv(bzfile(local_filename), strip.white=TRUE)


##Need to deal with EVTYPE



#convert the damage and their partner exponent vairables to a cash variable
#we will use the below qdap table to map exponent to a multiplier variable and then to cash

h = 10^2
k = 10^3
th = 10^4
hh = 10^5
m = 10^6
b = 10^9

key <- data.frame(
        c( "H", "h", "2", 
           "k", "K", "3", 
           "m", "M", "6", 
           "b", "B","9"),
        c( h, h, h,
           k, k, k,
           m, m, m,
           b, b, b)
)

colnames(key) <- c("key", "value")


storm_data$crop_multiplier <- lookup(key.match = key, storm_data$CROPDMGEXP, missing = 1)
storm_data$crop_cash <- storm_data$CROPDMG * storm_data$crop_mult

storm_data$prop_multiplier <- lookup(key.match = key, storm_data$PROPDMGEXP, missing = 1)
storm_data$prop_cash <- storm_data$PROPDMG * storm_data$prop_mult

##need to figure out inflation


## There are a lot of columns we can drop


```



##Results
###Is there a section titled "Results" where the main results are presented?


####summary tables
####two plots: econimc damage, bodily harm

###Is there at least one figure in the document that contains a plot?
###Are there at most 3 figures in this document?