---
title: "Identifing Environmental Events With Greatest Health and Economic Effects"
date: "`r Sys.Date()`"
output: html_document
keep_md: yes
---

Unzip and read source data. We will also load libraries that we will use later in analysis at this stage.


```{r initialize}

library(dplyr)
library(qdap)

source_url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
local_filename = "StormData.csv.bz2"

evtype_file = "official_evtype.csv"

writeLines(capture.output(sessionInfo()), "sessionInfo.txt")

```


##Synopsis
###Does the document have a synopsis that describes and summarizes the data analysis in less than 10 sentences?


This analysis is preformed on the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. Data was downloaded on `r Sys.Date()` from `r source_url`.



##Data Processing
###Is there a section titled "Data Processing" that describes how the data were loaded into R and processed for analysis?
Since NOAA only began recording all events since 1996, all data for prior years will be dropped. Source: http://www.ncdc.noaa.gov/stormevents/details.jsp?type=eventtype

A list of official event types was collected by hand from the ￼NATIONAL WEATHER SERVICE INSTRUCTION 10-1605 document published on August 17, 2007. At the time of this report, the document was available here:

http://www.nws.noaa.gov/directives/sym/pd01016005curr.pdf

The official events are saved in `r evtype_file`.


```{r get-and-clean-source-data}


#download.file(source_url, destfile = local_filename)

storm_data <- read.csv(bzfile(local_filename), strip.white=TRUE)


#Pull the year out of the BGN_DATE column.
regex_pattern <- "\\d\\d?/\\d\\d?/(\\d\\d\\d\\d).*"
storm_data$year <- as.numeric(gsub(regex_pattern, "\\1", storm_data$BGN_DATE))

#since we want to be fair to all the events
storm_data <- storm_data[storm_data$year >= "1996",]


##Need to deal with EVTYPE
evtypes <-read.csv("official_evtype.csv")
evtypes$EVTYPE_offical <- toupper(evtypes$EVTYPE_offical)


#convert the damage and their partner exponent vairables to a cash variable
#we will use the below qdap table to map exponent to a multiplier variable and then to cash

h = 10^2
k = 10^3
th = 10^4
hh = 10^5
m = 10^6
b = 10^9

key <- data.frame(
        c( "H", "h", "2", 
           "k", "K", "3", 
           "m", "M", "6", 
           "b", "B","9"),
        c( h, h, h,
           k, k, k,
           m, m, m,
           b, b, b)
)

colnames(key) <- c("key", "value")


storm_data$crop_multiplier <- lookup(key.match = key, storm_data$CROPDMGEXP, missing = 1)
storm_data$crop_cash <- storm_data$CROPDMG * storm_data$crop_mult

storm_data$prop_multiplier <- lookup(key.match = key, storm_data$PROPDMGEXP, missing = 1)
storm_data$prop_cash <- storm_data$PROPDMG * storm_data$prop_mult


## There are a lot of columns we can drop


```



##Results
###Is there a section titled "Results" where the main results are presented?


####summary tables
####two plots: econimc damage, bodily harm

###Is there at least one figure in the document that contains a plot?
###Are there at most 3 figures in this document?