---
title: "Identifing Environmental Events With Greatest Health and Economic Effects"
auhtor: "Philip Tabak"
date: "`r Sys.Date()`"
output: html_document
keep_md: yes
---

Unzip and read source data. We will also load libraries that we will use later in analysis at this stage.


```{r initialize}

library(dplyr)

source_url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
local_filename = "StormData.csv.bz2"

writeLines(capture.output(sessionInfo()), "sessionInfo.txt")

#This mapping function will be used to tidy monetary units
calcDollars <- function(num, code){
        
        num <= as.double(num)
        return_val <- as.double(num)
        code <- toupper(code)
        
        if(code == "B"){return_val = num * 1000000000}
        if(code == "M"){return_val = num * 1000000}
        if(code == "K"){return_val = num * 1000}
        if(code == "H"){return_val = num * 100}
        
        return(return_val)
}
```


##Synopsis
###Does the document have a synopsis that describes and summarizes the data analysis in less than 10 sentences?


This analysis is preformed on the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. Data was downloaded on `r Sys.Date()` from `r source_url`.



##Data Processing
###Is there a section titled "Data Processing" that describes how the data were loaded into R and processed for analysis?


```{r get-and-clean-source-data}


download.file(source_url, destfile = local_filename)

storm_data <- read.csv(bzfile(local_filename))

storm_data_tidy <- mutate(storm_data, prop_dollars = calcDollars(PROPDMG, PROPDMGEXP))

```



##Results
###Is there a section titled "Results" where the main results are presented?


###Is there at least one figure in the document that contains a plot?
###Are there at most 3 figures in this document?