---
title: "Identifing Environmental Events With Greatest Health and Economic Effects"
date: "`r Sys.Date()`"
output: html_document
keep_md: yes
---

Unzip and read source data. We will also load libraries that we will use later in analysis at this stage.


```{r initialize}

library(dplyr)
library(qdap)
library(stringr)
library(stringdist)

source_url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
local_filename = "StormData.csv.bz2"

evtype_file = "official_evtype.csv"

writeLines(capture.output(sessionInfo()), "sessionInfo.txt")

```


##Synopsis
###Does the document have a synopsis that describes and summarizes the data analysis in less than 10 sentences?


This analysis is preformed on the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. Data was downloaded on `r Sys.Date()` from `r source_url`.



##Data Processing
###Is there a section titled "Data Processing" that describes how the data were loaded into R and processed for analysis?
Since NOAA only began recording all events since 1996, all data for prior years will be dropped. Source: http://www.ncdc.noaa.gov/stormevents/details.jsp?type=eventtype

A list of official event types was collected by hand from the ￼NATIONAL WEATHER SERVICE INSTRUCTION 10-1605 document published on August 17, 2007. At the time of this report, the document was available here:

http://www.nws.noaa.gov/directives/sym/pd01016005curr.pdf

The official events are saved in `r evtype_file`.


```{r get-and-clean-source-data}

download.file(source_url, destfile = local_filename)

storm_data_full <- read.csv(bzfile(local_filename), strip.white=TRUE)
#storm_data_full <- read.csv(local_filename, strip.white=TRUE)

cols_to_keep <- c("BGN_DATE", 
                        "CROPDMG",
                        "CROPDMGEXP",
                        "PROPDMG",
                        "PROPDMGEXP",
                        "EVTYPE",
                        "INJURIES",
                        "FATALITIES")

storm_data <- storm_data_full[cols_to_keep]

#Pull the year out of the BGN_DATE column.
regex_pattern <- "\\d\\d?/\\d\\d?/(\\d\\d\\d\\d).*"
storm_data$YEAR <- as.numeric(gsub(regex_pattern, "\\1", storm_data$BGN_DATE))

#since we want to be fair to all the events
storm_data <- storm_data[storm_data$YEAR >= "1996",]

```


####EVTYPE needs a lot of work.
```{r EVTYPE-cleanup}
evtypes <- read.csv("official_evtype.csv", strip.white = TRUE)
evtypes$EVTYPE.offical <- toupper(evtypes$EVTYPE.offical)

#convert all of EVTYPE to upper case
storm_data$EVTYPE.clean <- str_trim(toupper(storm_data$EVTYPE))

#remove all summary rows, i.e., rows containing the word "SUMMARY" from EVTYPE.clean
storm_data <- storm_data[grep(".*(SUMMARY).*$", storm_data$EVTYPE.clean, invert = TRUE),]

#replace all instances of TSTM with THUNDERSTORM
storm_data$EVTYPE.clean <- gsub("TSTM", "THUNDERSTORM", storm_data$EVTYPE.clean, ignore.case = TRUE)
#cold should always be followed by /wind chill
storm_data$EVTYPE.clean <- gsub("COLD$", "COLD/WIND CHILL", storm_data$EVTYPE.clean, ignore.case = TRUE)
#hurricane should always be followed by /typhon
storm_data$EVTYPE.clean <- gsub("HURRICANE$", "HURRICANE/TYPHOON", storm_data$EVTYPE.clean, ignore.case = TRUE)
#frost should always be followed by /freeze
storm_data$EVTYPE.clean <- gsub("FROST$", "FROST/FREEZE", storm_data$EVTYPE.clean, ignore.case = TRUE)
#get rid of "DAMAGE"
storm_data$EVTYPE.clean <- gsub(" DAMAGE", "", storm_data$EVTYPE.clean, ignore.case = TRUE)
#take out any brackets and their contents
storm_data$EVTYPE.clean <- gsub("\\(.*\\)", "", storm_data$EVTYPE.clean, ignore.case = TRUE)
#if we've managed to add leading and trailing whitesapce, get rid of that
storm_data$EVTYPE.clean <- str_trim(storm_data$EVTYPE.clean)
#take out any trailing Ss. The official does not contain pluralizations
storm_data$EVTYPE.clean <- gsub("S$","", storm_data$EVTYPE.clean, ignore.case = TRUE)
#take out any double spaces
storm_data$EVTYPE.clean <- gsub("  ", " ", storm_data$EVTYPE.clean)
#i think it's possible we introduced whitespace, better trim again
storm_data$EVTYPE.clean <- str_trim(storm_data$EVTYPE.clean)


match_vector <- amatch(storm_data$EVTYPE.clean , evtypes$EVTYPE.offical)
storm_data$EVTYPE.clean <- evtypes$EVTYPE.offical[match_vector]


#finally, anything that we haven't converted should go from NA to OTHER

storm_data$EVTYPE.clean[is.na(storm_data$EVTYPE.clean)] <- "OTHER"
```

```{r convert-to-cash}
#convert the damage and their partner exponent vairables to a cash variable
#we will use the below qdap table to map exponent to a multiplier variable and then to cash

h = 10^2
k = 10^3
th = 10^4
hh = 10^5
m = 10^6
b = 10^9

key <- data.frame(
        c( "H", "h", "2", 
           "k", "K", "3",
           "4",
           "5",
           "m", "M", "6", 
           "b", "B","9"),
        c( h, h, h,
           k, k, k,
           th,
           hh,
           m, m, m,
           b, b, b)
)

colnames(key) <- c("key", "value")

#perform a simple key match to get a multiplier, then multiply to cash
storm_data$CROP.multiplier <- lookup(key.match = key, storm_data$CROPDMGEXP, missing = 1)
storm_data$CROP_DAMAGE <- storm_data$CROPDMG * storm_data$CROP.multiplier

#repeate for property
storm_data$PROP.multiplier <- lookup(key.match = key, storm_data$PROPDMGEXP, missing = 1)
storm_data$PROPERTY_DAMAGE<- storm_data$PROPDMG * storm_data$PROP.multiplier

```

##get rid of intermediary veriables we are not using
```{r cleanup}
# get rid of some of the working columns, just keep what he need going forward
cols_to_keep <- c("YEAR", 
                  "EVTYPE.clean",
                  "INJURIES",
                  "FATALITIES",
                  "CROP_DAMAGE",
                  "PROPERTY_DAMAGE")
                
storm_data <- storm_data[cols_to_keep]
```



##Results
###Is there a section titled "Results" where the main results are presented?


####summary tables
####two plots: econimc damage, bodily harm

###Is there at least one figure in the document that contains a plot?
###Are there at most 3 figures in this document?